Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-17 paper/03-methods.qmd
Big up methods and results

Social mobility effects—understood as impacts on an outcome arising from movements between an origin state and a destination state (e.g., social class)—have long been a focus of sociological research [@eyles_social_2022; @langsaether_explaining_2022]. Yet, as Breen and Ermisch [-@breen_effects_2024, p.467] emphasize, most of the mobility hypotheses are, at their core, individual-level counterfactual comparisons between the observed outcome under a mobility trajectory and the outcome the same person would have shown if they had remained in their origin class (or moved to an alternative destination). Standard specifications (e.g., SAM, DRM, and mobility-contrast models) struggle to retrieve these inherently counterfactual quantities due to identification constraints and their reliance on between-group contrasts constructed from additive terms and interactions, thereby yielding primarily associational evidence [@breen_effects_2024; @song_there_2025]. I therefore adopt Breen and Ermisch’s [-@breen_effects_2024] causal formulation, treating destination class as the treatment, conditioning on the class of origin, and estimating heterogeneous destination effects with observational data under explicit identification assumptions. To align design and target, I follow the MIDA template [@blair_research_2023]: set out the causal model and assumptions (M), define the inquiry and estimand (I), describe the data and variables (D), and detail the answer strategy (A) used to identify the causal effect of intergenerational social mobility on preferences for the commodification of pensions.

@@

Breen and Ermisch [-@breen_effects_2024], building on a critical reassessment of the inferential limits of standard mobility models, propose a causal framework that reconceptualizes the “mobility effect” as the treatment effect of reaching a destination class, with impacts heterogeneous by class of origin. The core claim is that typical mobility hypotheses are fundamentally within-person counterfactual comparisons rather than between-person contrasts. Defining mobility as $M = D - O$ —a change from origin $(O)$ to destination $(D)$, rather than an additive or interactive combination—places the focus squarely on the Neyman–Rubin problem: for any individual, we observe the result in the social position they currently occupy, but not in the situation of other alternative destinations or immobility.

@@

In fact, establishing the causal relationship between objective social mobility and subjective outcomes, such as preferences, poses several challenges for causal inference. Below, I use directed acyclic graphs (DAGs) to illustrate some of these challenges and evaluate possible strategies for identifying the effects of social mobility. @fig-dag depicts the causal model guiding identification. In this model, a person's preference ($Y$) is directly caused by her class destination ($D$) and can also be affected by different attributes determined in childhood or earlier ($C$), such as family resources, household composition, and ascriptive status (pre-treatment confounders). Class origin ($O$) is determined by those origin-attributes ($C$) and, in turn, affects destination ($D$). In addition, unobserved factors ($U$) can also influence origin-attributes ($C$). There is no direct arrow from $O$ to $Y$, reflecting the assumption that the effect of origin on the outcome operates entirely through background factors $(C)$ and destination $(D)$.  Accordingly, conditioning on $(O,C)$ could blocks the relevant back-door paths from $D$ to $Y$, specifically $D\leftarrow O\leftarrow C\rightarrow Y$ and, if present, $D \leftarrow O \rightarrow Y$, so $(O,C)$ is a minimal sufficient adjustment set.

@@

Formally, the DAG encodes the following relations:

$$
U \rightarrow C,\quad C \rightarrow O,\quad C \rightarrow D,\quad C \rightarrow Y,\quad O \rightarrow D,\quad D \rightarrow Y
$$

This structure corresponds to a nonparametric structural causal model in which $D$ is the treatment variable, $Y$ the outcome, and $(O, C)$ the minimal sufficient adjustment set. Conditioning on these variables satisfies the back-door criterion for identifying the causal effect of mobility on preferences [@breen_effects_2024].

Why are the assumptions plausible here? Conditional independence is targeted by controlling for a plausible sufficient adjustment set $(O,C)$; inverse probability weights (IPW) obtained via entropy balancing enforces covariate balance within each origin stratum. Positivity is assessed via propensity scores and trimming of extreme weights, restricting to common support. SUTVA holds by treating the destination class as a single exposure measured before $(Y)$ and assuming no interference. Together, the DAG and assumptions define conditions for identifying the causal effect of intergenerational social mobility on preferences for commodification.

@@

The causal inquiry guiding this study asks: How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile? Following the potential outcomes framework proposed by Breen and Ermisch [-@breen_effects_2024, p.473], the estimand of interest is the average treatment effect on the treated (ATT), defined as the mean causal effect of moving from an origin class $(O=j)$ to a destination class $(D=k)$, compared with the counterfactual outcome that those same individuals would have exhibited had they instead reached an alternative destination $(k')$:

$$ 
ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
$$ {#eq-breen}

In this study, I focus on the specific case where the counterfactual destination corresponds to class immobility $(k' = j)$, that is, how the outcome among people from origin *j* who entered destination *k* would have bee different if those people had, counterfactually, remained in origin *j* instead (inmobility). Accordingly, the estimand becomes:

$$ 
ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
$$ {#eq-att}

which represents the average causal effect of attaining class *(k)*—for instance, moving upward or downward—relative to remaining in class *(j)*, among individuals who actually experienced such mobility. This estimand captures the within-origin counterfactual contrast at the heart of mobility effects: how a person’s preference for market-based pensions would differ if, counterfactually, they had stayed in their origin class rather than moved to their observed destination.

@@

Because respondent occupation is not measured in every wave, I restrict the analysis to the 2016, 2018, and 2023 waves. Parental occupation was collected only in 2023 as open-text labels. I coded these texts to the ISCO-08 two-digit scheme using the National Institute of Statistics of Chile’s automated coding API and retained cases with high-confidence matches (>95%). Treating parental occupation as a time-invariant origin attribute, I carried these codes back to 2016 and 2018. After listwise deletion and restricting to key variables (respondent occupation and the outcome measure), the final analytic sample comprises *N* = 3,435 observations nested within *N* = 1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with the study design, estimation proceeds within trajectory-specific subsamples (e.g., low→low; low→middle), retaining movers and matched non-movers from the same origin class. Then, I apply weights to construct the corresponding counterfactual control sample for each trajectory. Following Breen and Ermisch [-@breen_effects_2024], I use the three waves in the estimations that follow, allowing for correlation between the observations for each individual in calculating the standard errors (i.e. clustering on the personal identifier).

@@

The outcome variable measures preferences regarding the commodification of pensions, operationalized with a single item addressing how strongly individuals justify conditioning access to old-age pension benefits based on individual income. Respondents were asked: “Is it fair in Chile that people with higher incomes have better pensions than people with lower incomes?” and were asked to indicate their level of agreement on a five-point Likert scale ranging from 1 (“strongly disagree”) to 5 (“strongly agree”). For estimation and interpretability, I dichotomized the item so that values 4-5 indicate agreement with market-based access to old-age pensions (coded as 1), while values 1-3 indicate disagreement (coded as 0). This binary coding treats neutrality as substantially closer to non-support than to support and produces a clear contrast between respondents who support market-based differentiation in pensions and those who do not. I use this item for two main reasons: on the one hand, to be able to make comparisons with existing literature [@lindh_public_2015; @otero_power_2024; @castillo_perceptions_2025] and, on the other hand, because it captures two fundamental dimensions of market-based welfare distribution: the role of economic resources as a key determinant of outcomes and the framing of services as as tradable commodities that can be bought and sold based on ability to pay [@lindh_public_2015].

@tbl-summary1 shows the descriptive statistics of the variables used in the study.

```{r}
#| label: tbl-summary1
#| tbl-cap: "Descriptive statistics for all variables."
#| tbl-cap-location: top
#| results: asis
#| echo: false

t1 <- db %>% 
  select(just_pension_f, 
         estrato_orig, 
         estrato_ocupa, 
         educ_orig_f,
         hogar_bip_f, 
         nacionalidad_f, 
         sexo_f,
         m0_edad_f, 
         etnia_f, 
         merit_i,
         wave

)

t1 <- t1 %>% 
  mutate(
    just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree")),
    hogar_bip_f  = factor(hogar_bip_f, levels = c(0,1), labels = c("No co-residence", "Co-residence")),
    #nacionalidad_f = if_else(nacionalidad_f == 0, 1, 0),
    nacionalidad_f  = factor(nacionalidad_f, levels = c(0,1), labels = c("Non-Chilean", "Chilean")),
    #sexo_f = if_else(sexo_f == 1, 0, 1),
    sexo_f  = factor(sexo_f, levels = c(0,1), labels = c("Male", "Female")),
    etnia_f = factor(etnia_f, levels = c(0,1), labels = c("Non-indigenous", "Indigenous"))
  )

t1$just_pension_f <- sjlabelled::set_label(t1$just_pension_f, "Preference for pension commodification")
t1$educ_orig_f <- sjlabelled::set_label(t1$educ_orig_f, "Parental education")
t1$hogar_bip_f <- sjlabelled::set_label(t1$hogar_bip_f, "Co-residence with both parents at age 15")
t1$nacionalidad_f <- sjlabelled::set_label(t1$nacionalidad_f, "Nacionality")
t1$sexo_f <- sjlabelled::set_label(t1$sexo_f, "Sex")
t1$m0_edad_f <- sjlabelled::set_label(t1$m0_edad_f, "Age (in years)")
t1$etnia_f <- sjlabelled::set_label(t1$etnia_f, "Indigenous ethnicity")
t1$merit_i <- sjlabelled::set_label(t1$merit_i, "Meritocracy perception")
t1$wave <- sjlabelled::set_label(t1$wave, "Wave")

df<-dfSummary(t1,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

df$Variable <- NULL # delete variable column

print(df)

@@

I treat intergenerational occupational mobility as an exposure indicating whether respondents occupy a different socioeconomic status than their fathers, closely following Breen and Ermisch’s [-@breen_effects_2024] causal framework. Class assignment proceeds in two steps. First, I derive status for both origin (father) and destination (respondent) from two-digit ISCO-08 occupations using the International Socio-Economic Index of Occupational Status (ISEI). ISEI is a hierarchical, unidimensional scale that locates occupations by incumbents’ average education and earnings and was explicitly designed to harmonize occupational stratification across other class schemes and socioeconomic status measures [@ganzeboom_internationally_1996], thereby supporting cross-study comparability [@barozet_measurement_2021a]. Second, I group both origin and destination ISEI scores into terciles (low, middle, high), yielding a three-category schema for origin and destination stratums (see @tbl-matrix). Substantively, I use ISEI because occupation-based measures capture structural locations that organize life chances and attitudes more directly than income alone: they bundle access to the means of production, qualification credentials, and the authority/power inherent in labor relations; central dimensions of stratification and inequality [@barone_rise_2022; @wrightUnderstandingClass2015]. In short, using ISEI aligns the measurement of origin and destination on a common vertical status continuum that is theoretically meaningful for social mobility research and empirically feasible given the available information of ELSOC’s data.

@@

Within each origin stratum $O=j$, I estimate average treatment effects on the treated (ATT) by comparing movers to otherwise similar non-movers from the same origin. To adjust for selection into mobility, I reweight only the control group (immobile) using entropy balancing so that the distribution of pre-treatment origin covariates $C$ matches that of the treated on selected moments.

The selection of the adjustment set $C$ follows two principles. First, it mirrors the logic in Breen and Ermisch [-@breen_effects_2024]—who condition on attributes determined in childhood or earlier (e.g., parental resources, household structure, ascriptive traits)—and, second, it is constrained to origin-side characteristics available in ELSOC. Guided by theory and Chilean evidence on relative mobility determinants [mainly @brunori_inequality_2025; @celhay_intergenerational_2010; @espinoza_movilidad_2014;@torche_unequal_2005], I include six covariates covering family resources, household composition, and ascriptive status: (a) parental education (highest of father/mother), 10 categories coded ordinally from no schooling to postgraduate studies; (b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c) nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0 = male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes). These variables are fixed before the destination class, plausibly affect selection into mobility, and constitute the maximal origin set in ELSOC, thereby making the conditional independence assumption more credible within each $O=j$.

After balancing, I construct stabilized IPW, setting treated weights to 1 and assigning derived weights to controls, thus aligning the counterfactual distribution of non-movers with the covariate profile of movers in each origin stratum (see [Supplementary Material](#anexo) for the evaluation of balance). These weights are then used in the outcome stage to estimate the causal effect of mobility on preferences for the commodification of pensions.

@@

Because the dependent variable—the preference for pension commodification—is binary $(Y_{it}\in{0,1})$, I estimate weighted linear probability models (WLS) using stabilized inverse probability weights (IPW) obtained via entropy balancing. Although OLS estimator is often questioned with binary outcomes, it’s unbiased. It consistently estimates the conditional expectation $E(Y|X)$ under standard exogeneity $E[\varepsilon_i|X_i]=0$, with heteroskedasticity addressed by robust or clustered errors [@wooldridge_introductory_2009, ch. 7.5]. WLS further improves efficiency under heteroskedasticity or unequal sampling probabilities [@wooldridge_introductory_2009, ch. 8.4]. Following Gelman, Hill, and Vehtari [-@gelman_regression_2020, pp. 270–272], IPW-weighted regression transparently targets the ATT by recreating the counterfactual distribution. Linear models are preferred here because coefficients are average probability differences; nonlinear models (logit/probit) complicate weighting and interpretation [@gelman_regression_2020, ch. 13].

@@

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds if either the weighting model or the outcome model is correctly specified, while also improving efficiency [@wooldridge_introductory_2009]. These results are virtually identical to baseline models without covariates (see [Supplementary Material](#anexo) for complete models), with no substantive changes in the mobility coefficients. In addition, I assess causal heterogeneity by interacting the mobility treatment with (i) meritocratic perceptions and (ii) survey wave, reporting simple slopes (origin-specific ATTs) by moderator levels. To evaluate sensitivity to unmeasured confounding, I compute E-values for point estimates and confidence limits [@ding_sensitivity_2016]. To address multiplicity across the six mobility trajectories, I also report Holm-adjusted p-values alongside 95% confidence intervals.

All the analyses were conducted using R software and the _lm_robust_ package.

>>>>

Social mobility effects—understood as impacts on an outcome arising from movements between an origin state and a destination state (e.g., social class)—have long been a focus of sociological research [@eyles_social_2022; @langsaether_explaining_2022]. Yet, as Breen and Ermisch [-@breen_effects_2024, p.467] emphasize, most of the mobility hypotheses are, at their core, individual-level counterfactual comparisons between the observed outcome under a mobility trajectory and the outcome the same person would have shown if they had remained in their origin class (or moved to an alternative destination). Standard specifications (e.g., SAM, DRM, and mobility-contrast models) struggle to retrieve these inherently counterfactual quantities due to identification constraints and their reliance on between-group contrasts constructed from additive terms and interactions, thereby yielding primarily associational evidence [@breen_effects_2024; @song_there_2025]. Following the causal framework of Breen and Ermisch [-@breen_effects_2024], I conceive of the destination class as a treatment, conditioning on the origin class and estimating the heterogeneous effects of the destination with observational data under explicit identification assumptions. To align design and target, I follow the MIDA template [@blair_research_2023]: set out the causal model and assumptions (M), define the inquiry and estimand (I), describe the data and variables (D), and detail the answer strategy (A) used to identify the causal effect of intergenerational occupational mobility on preferences for the commodification of pensions.

@@

Breen and Ermisch [-@breen_effects_2024], building on a critical reassessment of the inferential limits of standard mobility models, propose a causal framework that reconceptualizes the “mobility effect” as the treatment effect of reaching a destination class, with impacts heterogeneous by class of origin. The core claim is that typical mobility hypotheses are fundamentally within-person counterfactual comparisons rather than between-person contrasts. Defining mobility as $M = D - O$; a change from origin $(O)$ to destination $(D)$, rather than an additive or interactive combination, places the focus squarely on the Neyman–Rubin problem: for any individual, we observe the result in the social position they currently occupy, but not in the situation of other alternative destinations or immobility.

@@

Establishing the causal relationship between objective social mobility and subjective outcomes, such as preferences, poses several challenges for causal inference. Below, I use directed acyclic graphs (DAGs) to illustrate some of these challenges and evaluate possible strategies for identifying the effects of social mobility. @fig-dag depicts the causal model guiding identification. In this model, a respondent’s preference $Y$ is directly affected by her class destination $D$, and indirectly influenced by a set of pre-treatment attributes $C$ rooted in childhood—family resources, household composition, parental education, and other ascriptive characteristics. Class origin $O$ is itself shaped by these background factors $C$ and, in turn, affects destination $D$. I allow for unobserved determinants $U$ that influence background factors $C$, but I assume that any such unobserved variation operates only through $C$. Thus, there are no direct paths $U \rightarrow D$ or $U \rightarrow Y$ beyond those mediated by $C$.

Under this structure, conditioning on $(O, C)$ blocks all relevant backdoor paths from $D$ to $Y$—notably $D \leftarrow O \leftarrow C \rightarrow Y$. This renders $(O, C)$ a minimal sufficient adjustment set for identifying the causal effect of mobility on preferences, in line with the framework proposed by @breen_effects_2024. Identification therefore relies on the assumption that background factors $C$ adequately summarize pre-treatment characteristics that jointly shape both mobility and attitudes.

@@

Why are the assumptions plausible here? Conditional independence is targeted by controlling for a plausible sufficient adjustment set $(O,C)$ and by using inverse probability weights obtained via entropy balancing [@hainmueller_entropy_2012], which enforce covariate balance within each origin stratum and mitigate selection on observables. Positivity is assessed by inspecting the distribution of the entropy-balancing weights and trimming observations with extreme weights, thus restricting inference to regions of common support. SUTVA holds by treating the destination class as a single exposure measured before $(Y)$ and assuming no interference. Together, the DAG and assumptions define conditions for identifying the causal effect of intergenerational social mobility on preferences for commodification.

@@

The causal inquiry guiding this study asks: How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile? Substantively, I am interested in the effect of mobility relative to immobility within a given class of origin.

Following the potential outcomes framework proposed by Breen and Ermisch [-@breen_effects_2024, p.473], the estimand of interest is an average treatment effect on the treated (ATT) defined within each origin class. I focus on the specific case in which the counterfactual destination corresponds to class immobility $(k' = j)$. In other words, I ask how the outcome among people from origin *j* who entered destination *k* would have bee different if those people had, counterfactually, remained in origin *j* instead (inmobility).

Formally,  for individuals with origin $(O=j)$ who attain destination $(D=k)$, the estimand is:

$$

ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]

$$ {#eq-att}

which represents the mean causal effect of moving from origin class $(j)$ to destination class $(k)$, compared to the counterfactual outcome those same individuals would have exhibited had they instead remained in $(j)$. Thus, this ATT captures the counterfactual contrast within origin that lies at the heart of mobility effects: how preferences for market-based pensions would differ if, for the same upwardly or downwardly mobile individuals, their observed destination class were replaced by immobility in their origin class.

@@

Because respondent occupation[^1] is not measured in every wave, I restrict the analysis to the 2016, 2018, and 2023 waves. After listwise deletion and restricting to key variables (respondent occupation and the outcome measure), the final analytic sample comprises *N* = 3,435 observations nested within *N* = 1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with the study design, estimation proceeds within trajectory-specific subsamples (e.g., low→low; low→middle), retaining movers and matched non-movers from the same origin class. Then, I apply weights to construct the corresponding counterfactual control sample for each trajectory. Following Breen and Ermisch [-@breen_effects_2024], I use the three waves in the estimations that follow, allowing for correlation between the observations for each individual in calculating the standard errors (i.e. clustering on the personal identifier).

[^1]: Parental occupation was collected only in 2023 as open-text labels. I coded these texts to the ISCO-08 two-digit scheme using the National Institute of Statistics of Chile’s automated coding API and retained cases with high-confidence matches (>95%). Treating parental occupation as a time-invariant origin attribute, I carried these codes back to 2016 and 2018.

@@

The outcome variable measures preferences regarding the commodification of pensions, operationalized with a single item addressing how strongly individuals justify conditioning access to old-age pension benefits based on individual income. Respondents were asked: “Is it fair in Chile that people with higher incomes have better pensions than people with lower incomes?” and answered on a five-point Likert scale ranging from 1 (“strongly disagree”) to 5 (“strongly agree”). For estimation and interpretability, I construct a binary indicator in which values 4–5 indicate agreement with market-based access to old-age pensions (coded 1), while values 1–3 indicate non-agreement (coded 0). Substantively, the dependent variable is meant to capture *endorsement* of pension commodification: only responses in the upper tail of the scale represent an active, explicit justification of market-based differentiation, whereas neutrality corresponds to an absence of such endorsement and is therefore treated as closer to non-support than to support.[^2] This binary coding thus yields a clear contrast between respondents who *support* market-based differentiation in pensions and those who do not. I use this item for two main reasons: first, to enable comparisons with existing work on market-based justice in social policy [@lindh_public_2015; @otero_power_2024; @castillo_perceptions_2025]; and second, because it taps two core dimensions of market-based welfare distribution—(i) the centrality of economic resources as a criterion for allocating outcomes and (ii) the framing of pensions as tradable commodities that can be bought and sold according to ability to pay [@lindh_public_2015].

[^2]: A potential concern is that the neutral category may reflect genuine ambivalence rather than disagreement. To address this, I conduct robustness checks using alternative codings: (i) excluding neutral respondents from the analysis; (ii) combining the neutral and agreement categories (3–5 vs. 1–2); and (iii) estimating models with the original 5-point scale using linear specifications. The main mobility effects reported below are substantively unchanged across these specifications (see [Supplementary Material](#anexo)).

@@

I treat intergenerational occupational mobility as an exposure indicating whether respondents occupy a different occupational status than their fathers, closely following Breen and Ermisch’s [-@breen_effects_2024] causal framework. Occupational assignment proceeds in two steps. First, I derive occupational status for both origin (father) and destination (respondent) from two-digit ISCO-08 codes using the International Socio-Economic Index of Occupational Status (ISEI). Second, I group these ISEI scores into terciles (low, middle, high), yielding a three-category schema for both origin and destination strata (see @tbl-matrix). Substantively, I use ISEI because occupational status is a core indicator of stratification and a reliable mobility measure: it locates jobs on a hierarchical continuum defined by incumbents’ typical education level and earnings [@hauser_intergenerational_2010; @salgado_uplifting_2025], and thus approximates long-run socio-economic position more closely than volatile, single-year income, which is also prone to recall and reporting error for both parental and own resources [@barone_rise_2022]. This occupational focus is consistent with a long tradition that interprets the occupational structure as the backbone of the stratification system and a key determinant of life chances [@wrightUnderstandingClass2015]. Moreover, ISEI was explicitly designed to harmonize occupational stratification across class schemes and SES measures [@ganzeboom_internationally_1996], supporting cross-study comparability. In short, using ISEI aligns the measurement of origin and destination on a common vertical status continuum that is theoretically meaningful for social mobility research and empirically feasible given the available information of ELSOC’s data.

@@

Within each origin stratum ($O = j$), I estimate average treatment effects on the treated (ATT) by comparing movers to otherwise similar non-movers from the same origin. To reduce bias from non-random selection into mobility based on observed characteristics, I preprocess the data using entropy balancing, reweighting only the control group (the immobile) so that the distribution of pre-treatment origin covariates $C$ among non-movers matches that of movers on selected moments. Entropy balancing implements a maximum-entropy reweighting scheme that chooses unit weights for non-movers to satisfy a set of balance constraints (e.g., equality of means and, where relevant, higher moments), while keeping the new weights as close as possible to the original weights [@hainmueller_entropy_2012]. Substantively, this procedure addresses selection on observables: it reweights immobile respondents so that, within each origin class, they resemble mobile respondents in their observed background characteristics, making any remaining differences in the outcome more plausibly attributable to mobility rather than to pre-existing measured advantages.

The selection of the adjustment set $C$ follows two principles. First, it mirrors the logic in Breen and Ermisch [-@breen_effects_2024], who condition on attributes determined in childhood or earlier (e.g., parental resources, household structure, ascriptive traits). Second, it is constrained to origin-side characteristics available in ELSOC. Guided by theory and Chilean evidence on the determinants of relative mobility [mainly @brunori_inequality_2025; @salgado_uplifting_2025; @espinoza_movilidad_2014; @torche_unequal_2005], I include six covariates covering family resources, household composition, and ascriptive status: (a) parental education (highest of father/mother), in 10 ordinal categories from no schooling to postgraduate studies; (b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c) nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0 = male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes). This set of covariates is fixed prior to the destination class and constitutes the maximum set on the origin side in ELSOC, thus allowing for the plausible capture of both family and contextual influences that may affect career trajectories, making the conditional independence hypothesis more credible within each $O = j$. Descriptive statistics for all variables are presented in Supplementary Material Table A1.

After achieving balance, I construct stabilized IPW-ATT weights by setting treated (mobile) cases to weight 1 and assigning the entropy-balancing weights to controls, rescaled within each origin stratum so that the mean weight remains close to one. This aligns the counterfactual distribution of non-movers with the covariate profile of movers in each origin class (see [Supplementary Material](#anexo) for balance diagnostics). The resulting weights are then used in the outcome stage to estimate the causal effect of intergenerational mobility on preferences for the commodification of pensions.

@@

Because the dependent variable—the preference for pension commodification—is binary $(Y_{it}\in{0,1})$, I estimate weighted linear probability models (WLS) using stabilized inverse probability weights (IPW) obtained via entropy balancing. Although OLS is often questioned with binary outcomes, it consistently estimates the conditional mean $E(Y|X)$ under standard exogeneity $E[\varepsilon_i|X_i]=0$, and heteroskedasticity can be handled with robust or clustered standard errors [@wooldridge_introductory_2009, ch. 7.5]. WLS with IPW further improves efficiency and implements the ATT estimand by recreating the counterfactual distribution of non-movers [@gelman_regression_2020, pp. 270–272]. Linear models are preferred here because coefficients are directly interpretable as average differences in predicted probabilities, whereas non-linear models complicate both weighting and interpretation [@gelman_regression_2020, ch. 13].

@@

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds if either the weighting model or the outcome model is correctly specified, while also improving efficiency [@wooldridge_introductory_2009]. These results are virtually identical to baseline models without covariates (see [Supplementary Material](#anexo) for complete models), with no substantive changes in the mobility coefficients. Heterogeneous effects by meritocratic perceptions and survey wave are examined through interaction models described in the previous section.

To evaluate the robustness of the findings, I implement two sets of sensitivity analyses. First, I conduct standard robustness checks by re-estimating the models under alternative codings of the outcome: (i) excluding neutral responses from the analysis, (ii) combining neutral responses with agreement (3–5 vs. 1–2), and (iii) using the original 5-point item in a linear specification. Second, I assess sensitivity to unmeasured confounding using the omitted-variable–bias framework of Cinelli and Hazlett [-@cinelli_making_2020], which quantifies how strong an unobserved confounder would need to be to explain away the estimated ATT.

### 2025-11-17 paper/04-analysis.qmd
Big up methods and results

## Models

@@

@tbl-inttime shows the causal heterogeneity of the mobility effect across the different waves of the survey. The treatment effect in 2016 (reference category) closely replicates the reference ATT patterns: positive for Middle→High ($\beta$ = 0.13, 95% CI [0.05, 0.22]) and negative for High→Middle ($\beta$ = −0.19, [−0.33, −0.05]) and High→Low ($\beta$ = −0.19, [−0.32, −0.05]). This indicates that, holding all other predictors constant, individuals who moved from middle to high socioeconomic status in 2016 were, on average, 13 percentage points more likely to support market-based access to the pension benefits compared to their counterfactual of immobility, while those who moved down from higher status were approximately 19 percentage points less likely. Beyond these reference effects, there is a notable temporal change in the High→Middle trajectory: the interaction between mobility and the 2023 wave is positive and statistically significant ($\beta$ = 0.19, [0.01, 0.38]). This suggests that, although people with downward mobility from high backgrounds were less supportive of commodification in 2016, by 2023 the same trajectory shows a change in sign, implying that the negative causal effect of downward mobility attenuated and eventually turned positive. No comparable changes are observed in other trajectories. In essence, this pattern reveals time-conditional heterogeneity in the mobility effect: among those who experienced downward mobility, preferences for market-based access to old-age pensions appear to have realigned in the most recent wave, reflecting a notable shift toward greater support for commodification within this trajectory by 2023.

## Sensitivity analysis

The first sensitivity check corresponds to the @ding_sensitivity_2016 approach to assess unmeasured confounding (E values). This values indicates the minimum strength of association that an unobserved confounder would need to have with both the treatment and the outcome—on the relative risk scale—to explain an effect. The Middle-High ATT would require an $RR ≈ 1.74$ to offset the point estimate ($E_{point}$ = 1.736) and an $RR ≈ 1.26$ to move its CI to include the null value ($E_{CI}$ = 1.264). High→Low shows comparable robustness ($E_{point}$ = 1.759; $E_{CI}$ = 1.239), while High→Middle is comparatively fragile near zero ($E_{point}$ = 1.591; $E_{CI}$ = 1.037). For non-significant trajectories (Low→Middle, Low→High, Middle→Low), $E_{CI}$ = 1, as expected. In conclusion, the directional pattern—stronger in favor of commodification in Middle→High and weaker in the descent from High—seems reasonably robust to moderate unmeasured confounding in Middle→High and High→Low, but borderline in High→Middle. These results justify treating signs and relative magnitudes as informative, while recognizing limited protection against omitted variables for High→Middle (see [Supplementary Material](#anexo) for complete tables).

The second sensitivity check applies Holm’s step-down correction, which is a multiple-testing procedure that controls the familywise error rate (FWER) while retaining more power than Bonferroni, across the six mobility trajectories. After FWER control, none of the effects remain significant at $\alpha$ = .05: Middle→High ($p_{raw}$ = .0116 → $p_{Holm}$ = .0698), High→Low (.0167 → .0834), and High→Middle (.0481 → .1924). Accordingly, claims of statistical significance should be tempered under strict multiplicity control. Substantively, however, this adjustment affects the strength of inference, not its direction: the pattern remains consistent with the theorized asymmetry (positive for Middle→High; negative for downward from high). I therefore treat signs and effect sizes as the primary evidence and regard p-values as exploratory once multiplicity is accounted for (see [Supplementary Material](#anexo) for complete tables).

>>>>

## Mobility effects models

@@

@tbl-inttime shows the causal heterogeneity of the mobility effect across the different waves of the survey. Regarding temporal heterogeneity, the only statistically significant interaction term is observed for the High→Middle trajectory in 2023 ($\beta_{\text{int}} = 0.19$), 95% CI [0.01, 0.38]). However, when this interaction is added to the baseline effect in 2016 (-0.19), the marginal effect of High→Middle mobility in 2023 is approximately zero, and its confidence interval includes the null. Thus, rather than indicating a reversal toward a positive effect, the evidence suggests that the negative impact of downward mobility from high backgrounds is confined to 2016 and attenuates over time. No comparable temporal changes are observed in the remaining trajectories. Overall, the results provide little support for strong time-conditional heterogeneity in mobility effects, beyond the fading of the initially negative High→Middle effect.

## Robustness check and sensitivity analysis

As a robustness check, I re-estimate the models under three alternative codings of the outcome: (i) recoding the neutral category as agreement (3–5 vs. 1–2), (ii) excluding neutral responses from the analysis, and (iii) using the original 5-point item in a linear specification. Across these alternative models (see [Supplementary Material](#anexo) for complete tables), the substantive pattern of the main effects is preserved: the Middle→High trajectory remains positively associated with support for pension commodification, and the High→Low trajectory remains negatively associated with it, with both coefficients generally increasing in magnitude. By contrast, the negative High→Middle effect becomes statistically indistinguishable from zero in the robustness specifications, indicating that this particular estimate is more sensitive to how the outcome is coded than the other two trajectories.

To assess the extent to which the estimated mobility effects may be driven by unmeasured confounding, I implement the sensitivity analysis proposed by Cinelli and Hazlett [-@cinelli_making_2020], which quantifies how strong an unobserved confounder would need to be—in terms of its joint explanatory power for both treatment and outcome—to attenuate or explain away the ATT, using gender as a benchmark covariate (the predictor with the largest partial association with the outcome aside from treatment). For the Middle→High trajectory ($\beta \approx 0.09$), partial $R^2 \approx 0.013$, an unobserved confounder would need to explain about 10–11% of the residual variance of both treatment and outcome to reduce the effect to zero and roughly 4% to render it statistically insignificant, substantially more than gender does. For the High→Low effect ($\beta \approx -0.10$), partial $R^2 \approx 0.012$, the required strength is of similar magnitude (around 10–11% to explain away the effect and about 4% to remove significance), again exceeding the explanatory power of the main observed covariates. By contrast, the High→Middle effect ($\beta \approx -0.07$), partial $R^2 \approx 0.007$ is somewhat less robust: a confounder explaining around 8% of residual variance in both treatment and outcome could nullify the estimate, and one explaining about 2% could make it non-significant. Overall, the sensitivity analysis suggests that the positive Middle→High and negative High→Low effects are moderately robust to unmeasured confounding, whereas the High→Middle effect is more vulnerable to relatively modest omitted variables (see [Supplementary Material](#anexo) for complete tables).
